Federated Learning's Blessing:\\ FedAvg has Linear Speedup

### 2020-06-04 source/sec_problem.tex
minor fix
Update sec_problem.tex

In this paper, we study the following federated learning problem:
\begin{align}
	\min _{\mathbf{w}}\left\{F(\mathbf{w}) \triangleq \sum\nolimits_{k=1}^{N} p_{k} F_{k}(\mathbf{w})\right\},
	\label{eq:problem}
\end{align}
where $N$ is the number of local devices (users/nodes/workers) and $p_k$ is the $k$-th device's weight satisfying $p_k \geq 0$ and $\sum_{k=1}^N p_k = 1$. 
In $k$-th local device, there are $n_k$ data points:
$\vx_k^1, \vx_k^2, \dots, \vx_k^{n_k}$.  
The local objective $F_k(\cdot)$ is defined as:
% \begin{align}
$F_{k}(\mathbf{w}) \triangleq \frac{1}{n_{k}} \sum_{j=1}^{n_{k}} \ell\left(\mathbf{w} ; \vx_k^j\right)$,
% \label{eq:localloss}
% \end{align}
where $\ell$ denotes a user-specified loss function. Each device only has access to its local data, which gives rise to its own local objective $F_k$. Note that we do not make any assumptions on the
data distributions of each local device. The local minimum
$\vw^{*}_k= \argmin_{\vw \in R^d} F_k(\vw)$ can be far from
the global minimum of \eq{\ref{eq:problem}}.

@@

\subsection{The Federated Averaging (FedAvg) Algorithm}
We first introduce the standard Federated Average (FedAvg) algorithm~\cite{mcmahan2016communication}.
FedAvg updates the model in each device by local Stochastic Gradient Descent (SGD) and send the latest model to the central server every $E$
steps. The central server conducts a weighted average over the model parameters
received from active devices and broadcasts the latest averaged model to all devices.
Formally, the updates of FedAvg at round $t$ is described as follows:
\begin{align}
\vv_{t+1}^{k} & =\vw_{t}^{k}-\alpha_{t}\vg_{t,k}\\
\mathbf{w}_{t+1}^{k} &=\left\{
\begin{array}{ll}
\mathbf{v}_{t+1}^{k} & \text { if } t+1 \notin \mathcal{I}_{E}, \\ 
\sum_{k \in \cS_{t+1}} \mathbf{v}_{t+1}^{k} & \text { if } t+1 \in \mathcal{I}_{E}
\end{array}\right.
\end{align}
where $\vw_t^k$ is the local model parameter maintained in the $k$-th device at the $t$-th iteration, $\vg_{t,k}:=\nabla F_{k}(\vw_{t}^{k},\xi_{t}^{k})$ is the stochastic gradient based on $\xi_{t}^{k}$, the data sampled from $k$-th device’s local data uniformly at random. $\mathcal{I}_{E}=\{E,2E,\dots\}$ is the set of global communication steps. We use $\mathcal{S}_{t+1}$ to represent the set of active devices at $t+1$.

Since federated learning usually involves an enormous amount of 
local devices, it is often more realistic to assume only a subset of 
local devices is active at each communication round (system heterogeneity). In this work,
we consider both the case of \textbf{full participation} where the model is
averaging over all devices at the communication round, i.e.,
$\mathbf{w}_{t+1}^{k} = \sum_{k=1}^N p_k \mathbf{v}_{t+1}^{k}$, and
the case of \textbf{partial participation} where $|\cS_{t+1}| < N$. 
With partial participation, $\cS_{t+1}$ is obtained by two types of
sampling schemes to simulate practical scenarios~\cite{li2019convergence}. For example, Sampling scheme I establishes $\cS_{t+1}$ by i.i.d. sampling the devices with probability $p_k$ with replacement. 
Both schemes
guarantee that gradient updates in FedAvg are unbiased stochastic versions of
updates in FedAvg with full participation. For 
more details on the notations and setup, please refer to Section~\ref{sec:app:notations} in the appendix.

@@

\begin{assumption}[L-smooth]
	$F_{1}, \cdots, F_{N}$ are all $L$-smooth: for all  $\mathbf{v}$  and $\mathbf{w}$, $F_{k}(\mathbf{v}) \leq F_{k}(\mathbf{w})+(\mathbf{v}- \\ \mathbf{w})^{T} \nabla F_{k}(\mathbf{w})+\frac{L}{2}\|\mathbf{v}-\mathbf{w}\|_{2}^{2}$.
	\label{ass:lsmooth}
\end{assumption}
\begin{assumption}[Strongly-convex]
	$	F_{1}, \cdots, F_{N} \text { are all } \mu \text { -strongly convex: for all v and } \mathbf{w}, F_{k}(\mathbf{v}) \geq F_{k}(\mathbf{w})+(\mathbf{v}-\mathbf{w})^{T} \nabla F_{k}(\mathbf{w})+\frac{\mu}{2}\|\mathbf{v}-\mathbf{w}\|_{2}^{2}$
	\label{ass:stroncvx}
\end{assumption}
\begin{assumption}[Bounded local variance]
	Let $\xi_{t}^{k}$ be sampled from the $k$-th device's local data uniformly at random. The variance of stochastic gradients in each device is bounded: $\mathbb{E}\left\|\nabla F_{k}\left(\mathbf{w}_{t}^{k}, \xi_{t}^{k}\right)-\nabla F_{k}\left(\mathbf{w}_{t}^{k}\right)\right\|^{2} \leq \sigma_{k}^{2}$,
	for $k=1, \cdots, N$ and any $\mathbf{w}_{t}^{k}$. Let $\sigma^2=\sum_{k=1}^{N}p_k\sigma_{k}^{2}$.
	\label{ass:boundedvariance}
\end{assumption}
\begin{assumption}[Bounded local gradient]
	The expected squared norm of stochastic gradients is uniformly bounded. i.e.,
	$\mathbb{E}\left\|\nabla F_{k}\left(\mathbf{w}_{t}^{k}, \xi_{t}^{k}\right)\right\|^{2} \leq G^{2}$, for all $k = 1,..., N$ and $t=0, \dots, T-1$.
	\label{ass:subgrad2}
\end{assumption}
Assumptions~\ref{ass:boundedvariance} and~\ref{ass:subgrad2} have been made in many previous
works in federated learning, e.g.~\cite{yu2019parallel,li2019convergence,stich2018local}. We provide some more justification for their generality. As model average parameters become
closer to $\mathbf{w}^{\ast}$, the $L$-smoothness property implies
that $\mathbb{E}\|\nabla F_{k}(\mathbf{w}_{t}^{k},\xi_{t}^{k})\|^{2}$
and $\mathbb{E}\|\nabla F_{k}(\mathbf{w}_{t}^{k},\xi_{t}^{k})-\nabla F_{k}(\mathbf{w}_{t}^{k})\|^{2}$
approach $\mathbb{E}\|\nabla F_{k}(\mathbf{w}^{\ast},\xi_{t}^{k})\|^{2}$
and $\mathbb{E}\|\nabla F_{k}(\mathbf{w}^{\ast},\xi_{t}^{k})-\nabla F_{k}(\mathbf{w}^{\ast})\|^{2}$.
Therefore, there is no substantial difference between these assumptions
and assuming the bounds at $\mathbf{w}^{\ast}$ only. Furthermore, compared to assuming \textit{bounded gradient diversity} as in related works~\cite{haddadpour2019convergence,li2018federated}, Assumption~\ref{ass:subgrad2} is much less restrictive. When target accuracy $\epsilon \rightarrow 0$,
bounded gradient diversity restricts local objectives to have the same minimizer as the global objective, contradicting the heterogeneous data setting. 
For detailed explanations of our assumptions, please refer to Appendix Section~\ref{sec:app:comparison}.

>>>>

In this paper, we study the following federated learning problem:
\begin{align}
	\min _{\mathbf{w}}\left\{F(\mathbf{w}) \triangleq \sum\nolimits_{k=1}^{N} p_{k} F_{k}(\mathbf{w})\right\},
	\label{eq:problem}
\end{align}
where $N$ is the number of local devices (users/nodes/workers) and $p_k$ is the $k$-th device's weight satisfying $p_k \geq 0$ and $\sum_{k=1}^N p_k = 1$. 
In the $k$-th local device, there are $n_k$ data points:
$\vx_k^1, \vx_k^2, \dots, \vx_k^{n_k}$.  
The local objective $F_k(\cdot)$ is defined as:
% \begin{align}
$F_{k}(\mathbf{w}) \triangleq \frac{1}{n_{k}} \sum_{j=1}^{n_{k}} \ell\left(\mathbf{w} ; \vx_k^j\right)$,
% \label{eq:localloss}
% \end{align}
where $\ell$ denotes a user-specified loss function. Each device only has access to its local data, which gives rise to its own local objective $F_k$. Note that we do not make any assumptions on the
data distributions of each local device. The local minimum
$F^{*}_k= \min_{\vw \in R^d} F_k(\vw)$ can be far from
the global minimum of \eq{\ref{eq:problem}}.

@@

\subsection{The Federated Averaging (FedAvg) Algorithm}
We first introduce the standard Federated Averaging (FedAvg) algorithm~\cite{mcmahan2016communication}.
FedAvg updates the model in each device by local Stochastic Gradient Descent (SGD) and send the latest model to the central server every $E$
steps. The central server conducts a weighted average over the model parameters
received from active devices and broadcasts the latest averaged model to all devices.
Formally, the updates of FedAvg at round $t$ is described as follows:
\begin{align}
\vv_{t+1}^{k} & =\vw_{t}^{k}-\alpha_{t}\vg_{t,k}\\
\mathbf{w}_{t+1}^{k} &=\left\{
\begin{array}{ll}
\mathbf{v}_{t+1}^{k} & \text { if } t+1 \notin \mathcal{I}_{E}, \\ 
\sum_{k \in \cS_{t+1}} \mathbf{v}_{t+1}^{k} & \text { if } t+1 \in \mathcal{I}_{E}
\end{array}\right.
\end{align}
where $\vw_t^k$ is the local model parameter maintained in the $k$-th device at the $t$-th iteration, $\vg_{t,k}:=\nabla F_{k}(\vw_{t}^{k},\xi_{t}^{k})$ is the stochastic gradient based on $\xi_{t}^{k}$, the data sampled from $k$-th device’s local data uniformly at random. $\mathcal{I}_{E}=\{E,2E,\dots\}$ is the set of global communication steps. We use $\mathcal{S}_{t+1}$ to represent the set of active devices at $t+1$.

Since federated learning usually involves an enormous amount of 
local devices, it is often more realistic to assume only a subset of 
local devices is active at each communication round (system heterogeneity). In this work,
we consider both the case of \textbf{full participation} where the model is
averaging over all devices at the communication round, i.e.,
$\mathbf{w}_{t+1}^{k} = \sum_{k=1}^N p_k \mathbf{v}_{t+1}^{k}$, and
the case of \textbf{partial participation} where $|\cS_{t+1}| < N$. 
With partial participation, $\cS_{t+1}$ is obtained by two types of
sampling schemes to simulate practical scenarios~\cite{li2019convergence}. For example, one scheme establishes $\cS_{t+1}$ by i.i.d. sampling the devices with probability $p_k$ with replacement. 
Both schemes
guarantee that gradient updates in FedAvg are unbiased stochastic versions of
updates in FedAvg with full participation. For 
more details on the notations and setup, please refer to Section~\ref{sec:app:notations} in the appendix.

@@

\begin{assumption}[L-smooth]
	$F_{1}, \cdots, F_{N}$ are all $L$-smooth: for all  $\mathbf{v}$  and $\mathbf{w}$, $F_{k}(\mathbf{v}) \leq F_{k}(\mathbf{w})+(\mathbf{v}- \\ \mathbf{w})^{T} \nabla F_{k}(\mathbf{w})+\frac{L}{2}\|\mathbf{v}-\mathbf{w}\|_{2}^{2}$.
	\label{ass:lsmooth}
\end{assumption}
\begin{assumption}[Strongly-convex]
	$	F_{1}, \cdots, F_{N} \text { are all } \mu \text { -strongly convex: for all v and } \mathbf{w}, F_{k}(\mathbf{v}) \geq F_{k}(\mathbf{w})+(\mathbf{v}-\mathbf{w})^{T} \nabla F_{k}(\mathbf{w})+\frac{\mu}{2}\|\mathbf{v}-\mathbf{w}\|_{2}^{2}$
	\label{ass:stroncvx}
\end{assumption}
\begin{assumption}[Bounded local variance]
	Let $\xi_{t}^{k}$ be sampled from the $k$-th device's local data uniformly at random. The variance of stochastic gradients in each device is bounded: $\mathbb{E}\left\|\nabla F_{k}\left(\mathbf{w}_{t}^{k}, \xi_{t}^{k}\right)-\nabla F_{k}\left(\mathbf{w}_{t}^{k}\right)\right\|^{2} \leq \sigma_{k}^{2}$,
	for $k=1, \cdots, N$ and any $\mathbf{w}_{t}^{k}$. Let $\sigma^2=\sum_{k=1}^{N}p_k\sigma_{k}^{2}$.
	\label{ass:boundedvariance}
\end{assumption}
\begin{assumption}[Bounded local gradient]
	The expected squared norm of stochastic gradients is uniformly bounded. i.e.,
	$\mathbb{E}\left\|\nabla F_{k}\left(\mathbf{w}_{t}^{k}, \xi_{t}^{k}\right)\right\|^{2} \leq G^{2}$, for all $k = 1,..., N$ and $t=0, \dots, T-1$.
	\label{ass:subgrad2}
\end{assumption}
Assumptions~\ref{ass:boundedvariance} and~\ref{ass:subgrad2} have been made in many previous
works in federated learning, e.g.~\cite{yu2019parallel,li2019convergence,stich2018local}. We provide further justification for their generality. As model average parameters become
closer to $\mathbf{w}^{\ast}$, the $L$-smoothness property implies
that $\mathbb{E}\|\nabla F_{k}(\mathbf{w}_{t}^{k},\xi_{t}^{k})\|^{2}$
and $\mathbb{E}\|\nabla F_{k}(\mathbf{w}_{t}^{k},\xi_{t}^{k})-\nabla F_{k}(\mathbf{w}_{t}^{k})\|^{2}$
approach $\mathbb{E}\|\nabla F_{k}(\mathbf{w}^{\ast},\xi_{t}^{k})\|^{2}$
and $\mathbb{E}\|\nabla F_{k}(\mathbf{w}^{\ast},\xi_{t}^{k})-\nabla F_{k}(\mathbf{w}^{\ast})\|^{2}$.
Therefore, there is no substantial difference between these assumptions
and assuming the bounds at $\mathbf{w}^{\ast}$ only. Furthermore, compared to assuming \textit{bounded gradient diversity} as in related works~\cite{haddadpour2019convergence,li2018federated}, Assumption~\ref{ass:subgrad2} is much less restrictive. When target accuracy $\epsilon \rightarrow 0$,
bounded gradient diversity restricts local objectives to have the same minimizer as the global objective, contradicting the heterogeneous data setting. 
For detailed explanations of our assumptions, please refer to Appendix Section~\ref{sec:app:comparison}.
